Statistical Theory\\ Chess Dataset Analysis

### 2024-09-29 Article/Article.tex
updated article

\section{Introduction}
The length of a chess match can be influenced by a variety of factors, including player skill level, strategy, and in-game dynamics. This study investigates the relationship between player ratings and the duration of chess matches, seeking to determine whether higher-rated players tend to play shorter or longer games. In addition to player ratings, we explore other game-related variables that might impact match length, such as opening moves and the type of result (win, draw, or loss).

Using a dataset of chess games, we apply statistical methods such as correlation analysis and regression models to analyze the influence of these factors. The study aims to offer insights into how skill level and game dynamics affect the length of a match, contributing to a broader understanding of player performance in chess.

>>>>

\section{Introduction}
Chess is one of humanity's oldest board games, played by two players on an $8\times8$ grid. Each player, controlling 16 pieces of their color, aims to checkmate the opponent's King, making it impossible for the King to escape capture. Unlike many games, chess does not involve luck or hidden information; the outcome is determined solely by the players' knowledge, strategy, and analytical skills.

The length of a chess match is influenced by several factors, including the players' skill levels, strategic choices, and in-game dynamics. In this project, we investigate the relationship between player ratings and the duration of chess matches, aiming to determine whether higher-rated players tend to play shorter or longer games. In addition to player ratings, we examine other variables that may impact match length, such as opening strategies, game outcomes (win, loss, draw), and time controls.

Our dataset comprises 19,113 games from Lichess, an online chess platform, each described by 16 features. The most relevant features to our analysis are: \begin{itemize} \item \textbf{Turns}: A positive integer representing the length of the game, where one turn includes a move by both White and Black. \item \textbf{ELO Rating}: A system that quantifies player skill, widely used by FIDE and online chess platforms, which we use to assess its correlation with match length. \item \textbf{Time Controls}: Competitive games are categorized into four main time control classes: Bullet, Blitz, Rapid, and Classical, based on initial clock times and increments. \item \textbf{Winner and Victory Status}: These columns capture whether White or Black won and the method of victory (checkmate, resignation, draw). \end{itemize} Through statistical analysis, we aim to understand how these factors influence the duration of chess matches and to provide insights into the relationship between skill level and game dynamics.
